Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study

### 2020-06-01 pharmaco.Rmd
updated pharmacoepi analyses

---
title: 'Associations of severe COVID-19 with drug prescribing in the REACT-SCOT case-control study'
author: 
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    email: paul.mckeigue@ed.ac.uk
    affiliation: HPS
  - name: Sharon Kennedy
    affiliation: HPS
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
csl: ../plos.csl
urlcolor: red

@@

_Methods_ -- Test positive COVID-19 cases ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database.  From the national population register ten controls per case were selected, matched for sex, age and general practice. Severe cases were those  who entered  a critical care unit or died within 28 days of testing.  International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary in the past nine months and used to define  history of listed conditions.  To examine if  any unexpected specific drug classes were likely to contribute to severe COVID19 risk, among  severe cases and controls without listed conditions  (INSERT n cases n controls) we undertook a backward selection procedure across all BNF chapters. Across the retained chapters we undertook a backward selection procedure at sub-paragraph level.   This yielded proton pump inhibitors (PPI) as the drug class that most explained the association of a prescription in the past year with COVID-19.  Accordingly we built a propensity score of PPI prescribing  across all 218 BNF subparagraph codes in a separate dataset of hospitalised non-severe cases and their controls.    This score, which increases as polypharmacy increases, was then used as a covariate that allowed adjustment of the association of PPIs with severe COVID-19 across all severe cases and their controls for confounding by co-prescribing. Rate ratios were estimated by conditional logistic regression.

*Findings* -- The risk of severe disease was strongly associated with the total number of drug classes dispensed in the last nine months.

Across sub-paragraphs of these chapters identified proton pump inhibitors as the drug class with the strongest independent association with severe COVID-19.  The rate ratio for severe COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[4, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 40 years. 
Adjustment for hospital diagnoses of disorders of the esophagus, stomach and duodenum reduced these effect sizes only slightly.  In those under 60 years only, the risks ratios increased with increasing average dispensed daily dose. More recent exposure was not more strongly associated than earlier exposure. With adjustment for propensity score associations were no longer significant in those aged 75+ but remained strong INSERT in those aged under 60 years. Restricting to those with no other dispensed prescriptions, the rate ratio associated with ever-use of a proton pump inhibitor was `r table.protonpump.bynumdrugs[1, 3]`.

*Conclusions* -- A systematic search for drug classes that underlie the association of severe COVID-19 with dispensing of any prescription has identified proton pump inhibitors as the drug class making the largest independent contribution to this association.  Prescription of this drug class is recognised as a a sentinel for polypharmacy and for frailty and  the  data were consistent with this  explanation for the association in the older population.   However the strength of the association in those under age 60, its independence of the propensity score and the observation of an association  in those not on any other drugs raise concern of a possible causal effect that requires further investigation.  In the meantime   existing advice on the avoidance of  inappropriate and over-prescribing of PPIS should be reinforced.

@@

# Background
The REACT-SCOT  study is a matched case control study of COVID-19 cases versus controls in Scotland.  In our first paper from this study we confirmed and quantified the risk ratios associated with conditions  such as diabetes, asthma and heart disease that had been  presumptively listed by the authorities as high risk for COVID-19 based on case series.      We further showed that in those without listed conditions prior hospitalisation in the previous five years  and prescription history  in the previous nine months  were strongly associated with severe disease.  We went on to show that the association with hospitalisation was not explained by just one or two other specific conditions but that many conditions of varying prevalence were more common in cases than controls.

A striking preliminary finding of the REACT-SCOT study was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year. The rate ratio associated with this variable varied from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]`

There are several potential explanations for such a finding;  there may be a sparse set of  drugs underlying this association or  there may be a large set of drugs contributing to it. These in turn may reflect  a i)sparse set of direct drug effects or ii) sparse set of underlying morbidities or a iii) larger set of direct drug effects  or  their interactions ( i.e. an effect due to the polypharmacy itself)  or a iv) larger set of co-morbidities ( i.e. that usually conceptualised as frailty) or even v) interactions between polypharmacy and frailty.

The objective of this study was to conduct an initial analysis to discover if there is a a sparse set of direct drug effects underlying the finding of the strong association of any prescription in the past year and severe COVID-19. Further analyses will explore the role of polypharmacy and frailty.

@@

## Case ascertainment and sampling of controls 
All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  For each test-positive case, the CHI database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the date of test in the case using the Community Health Index (CHI) database.  With this incidence density sampling design, odds ratios in conditional logistic regression models are equivalent to rate ratios.

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.

British National Formulary (BNF) drug codes were extracted from all prescriptions dispensed since 1 June 2019, excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.  The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).  BNF subparagraph codes generally represent pharmacological drug classes.  For this analysis prescription codes from chapters 14 and above, comprising mostly dressings and appliances but also vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each drug were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.  Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to 16 days before the date that the case (in a stratum consisting of case and matched controls) first tested positive.

# Statistical analysis 
As described previously, we derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  For this analysis the list was extended to include non-cardiac circulatory disease as our analyses of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that condition is the only indication.  Diagnoses of diabetes were extracted from linkage to the national diabetes register.  The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material.

@@

We constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.  To construct this score, we used an independent dataset so that the score would not overfit the case-control dataset.  As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, a dataset of `nrow(cc.hosp)` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls.  The propensity score was constructed on this training dataset by stepwise logistic regression with ever-use of a proton pump inhibitor as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription.

Rate ratios for severe COVID-19 were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.

# Results
## Relation of severe COVID-19 to number of drug classes dispensed. 
Table \ref{tab:numdrugs.agegr} shows that severe COVID-19 was strongly related to the total number of different drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed since 1 June 2019.  The strength of this association was present in all age groups. With no dispensed prescriptions as reference category, the rate ratios associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.

@@

As a next step, we partitioned the number of drug classes dispensed into totals for cardiovascular and non-cardiovascular drugs, on the basis that prescribing of multiple drugs for control of cardiovascular disease is supported by evidence-based guidelines, while prescribing of multiple drugs for other indications is more likely to represent inappropriate polypharmacy.  Table \ref{numdrugs,cardiovasc} shows that numbers of drugs in both categories were associated with COVID-19.

We next estimated univariate rate ratios for severe COVID-19 for indicator variables for dispensing of each of the `r length(subparacols)` BNF subparagraph codes.  Table XX shows the results for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed for at least 10 individuals and were associated with severe disease at $p < 0.001$.  These COVID-associated drug classes include many drug classes previously identified in Scotland as indicators of overprescribing and inappropriate polypharmacy.

## Backwards selection to identify specific drug classes contributing to the association of severe COVID-19 with number of drug classes dispensed
A backwards stepwise variable selection procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with number of drugs dispensed.  To save computation, the stepwise procedure was initiated with the the subset `r length(subparacols.forstepwisedrop)` drug classes that showed univariate associations.  The results of this procedure are shown in Table \ref{tab:stepwise.subparas}.  Dropping the first `r firstrow.neg - 1` drug classes from the set of drug classes used to compute the number of dispensed drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.

The only cardiovascular drug classes included among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.  Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.  The last drug class retained by backward selection, and thus the variable with largest independent explanatory power, was proton pump inhibitors.

## Associations of severe COVID-19 with proton pump inhibitor use
Testing for independent association of COVID-19 with exposure to proton pump inhibitors is challenging because these drugs are usually co-prescribed with other drugs, and because in this dataset the disease is strongly associated with polypharmacy.  We used several approaches:

* testing for a dose-response relationship,

* examination of the relationship by age.

* adjusting for prespecified covariates

* comparison of dispensing between recent and less time windows.

* construction of a propensity score using an independent dataset.

For these analyses the dataset was not restricted to exclude care home residents or those with diagnosed with a listed condition.

### Dose-response relationship
Table \ref{tab:dosegr.protonpump} shows how the effect size varies with the dose of proton pump inibitor, grouped into categories. Table \ref{tab:ppinames} compares the effects of the five drugs in this class.  Omeprazole and lansoprazole were by far the most commonly prescribed drugs.  The effect size, expressed as rate ratio for unit increase in dose expressed as average DDDs/day was higher for lansoprazole than for omeprazole.  This suggests that the DDDs for lansoprazole (30 mg) and omeprazole (20 mg) are not equivalent in potency.

### Relationship by age
Table \ref{tab:everuse.protonpump} shows the association with ever-use of proton pump inhibitors by age group, without and with adjustment for propensity score.  The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates.

Table \ref{tab:dose.protonpump} shows the rate ratio per unit increase in dose of proton pump inibitor, calculated as average defined daily doses dispensed per day since 1 June 2019.  There is a strong dose-response relationship in the two younger age groups, but not in those 
aged over 75 years.

### Adjusting for prespecified covariates

@@

## Construction of a propensity score
Table \ref{tab:numdrugs.protonpump} shows the association of ever-use of proton pump inhibitors with severe disease, stratified by number of other drugs dispensed.  This shows that the risk ratio associated with ever-use of proton pump inhibitors was XX in those who had no dispensed prescriptions of other drug classes.

As a next step, we constructed a propensity score for ever-use of a proton pump inhibitor. 
The `r nrow(coeffs.propensity)` variables retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.  As expected, the variables retained included NSAIDs and antiplatelet drugs, but also many other drug classes such as tricyclic analgesics that are not recognized medical indications for co-prescribing of a proton pump inhibitor. The coefficients for age and sex were dropped before calculating the propensity score in the case-control dataset, because the case-control analysis is matched for age and sex.

<!--- show effect of each drug class in propensity score before and after adjustment for PPI --->

## Associations with other drug classes
Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.  Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.

>>>>

---
title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study'
author: 
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    email: paul.mckeigue@ed.ac.uk
    affiliation: HPS
  - name: Sharon Kennedy
    affiliation: ISD
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart J McGurnaghan
    affiliation: IGMM
  - name: David McAllister\textsuperscript{\getAff{Glasgow}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}} 
    affiliation: HPS
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Thomas M Caparrotta 
    affiliation: IGMM
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Jim McMenamin
    affiliation: HPS
  - name: Colin Ramsay
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
address: 
  - code: Usher
    address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care

  - code: IGMM
    address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology.

  - code: HPS
    address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE

  - code: Glasgow
    address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Wellcome Trust Intermediate Clinical Fellow and Beit Fellow

  - code: Strathclyde
    address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology

  - code: GCU
    address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health

  - code: ISD
    address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health.

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
csl: ../plos.csl
urlcolor: red

@@

_Methods_ -- Severe COVID-19 cases were ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database and entered a critical care unit or died within 28 days of testing.  From the national population register ten controls per case were selected, matched for sex, age and general practice. International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary care during the past nine months. Rate ratios were estimated by conditional logistic regression.

*Findings* -- The drug classes associated with COVID-19 included widely-prescribed cardiovascular drugs and also multiple drug classes previously identified as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  This led us to examine associations with polypharmacy, defined as the total number of drug classes dispensed.  The rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.  The association with polypharmacy was strongest in those without past hospital admissions or diagnoses of conditions previously designated as conferring high risk.  A backwards elimination procedure identified proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of this association showed a dose-response relationship with the average daily dose dispensed and that the association was strongest in younger age groups: the rate ratio for severe COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.  The association remained after adjusting for covariates prespecified as likely confounders, and in those not prescribed any of the `r nrow(coeffs.propensity)` drug classes used to construct a propensity score.  However a comparison of time windows representing recent and less recent exposure in ever-users did not show a stronger association with recent exposure.

*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy. and this association is not easily explained by co-morbidity.  Of all drug classes, proton pump inhibitors make the largest independent contribution to this association.  As prescriptions of this drug class are a recognised indicator for overprescribing of drugs without clear medical indication, it is difficult to distinguish between causality and confounding as explanations for this association.  However the strength of the association in younger age groups and in those not on other drugs with which proton pump inhibitors are widely co-prescribed raise concern about a possible causal effect that requires further investigation.  In the meantime existing advice on reducing inappropriate polypharmacy and specifically limiting the prescribing of proton pump inhibitors should be reinforced.

@@

# Background
The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.  In the first report of this study we reported a strong association of severe COVID-19 with having had at least one dispensed prescription in the past year.  The univariate rate ratios associated with at least one dispensed prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.  This association persisted after adjusting for care home residence, any hospital admission in the last five yeasrs and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).  In a multivariate analysis adjusting for these covariates, the rate ratios five years or residence in a care home, the rate ratios associated with any dispensed prescription were 5.0 in those aged under 40 and 11 in those aged 75 years and over.  Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in almost all BNF chapters were associated with elevated rate ratios for COVID-19. In a multivariate analysis of all chapters, most of these associations were weaker, but prescriptions of at least one drug in chapters 1 (gastrointestinal system), 2 (nervous system) and 9 (nutrition and blood) remained strongly associated with COVID-19.  We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by the indications for which these drugs were prescribed or by frailty especially in older persons, causal effects of specific drugs or direct effects of polypharmacy on susceptibility could not be ruled out.

# Aim
The primary aim of this study was to explore the strong association of severe COVID-19 with dispensing of at least one prescription in the past year. Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.  A further aim was to determine whether tthe effects underlying this association are sparse or diffuse.  In the case of causal drug effects, this would correspond to distinguishing between large effects of a few drugs, or an effect of polypharmacy not attributable to any single drug.  In the case of confounding, this would correspond to distinguishing between identifying a few conditions that confer high-risk of COVID-19, or a diffuse effect of multimorbidity.

# Background
The REACT-SCOT study is a matched case-control study of severe COVID-19 in Scotland.  In the first report from this study an unexpected result was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.  The rate ratio associated with this variable varied from 10 in those aged under 60 years to 40
in those aged 75 years and over. The objective of this study was to establish how much of this association could be explained by confounding by co-morbidity and other health-related factors that influece drug prescribing and independently affect susceptibility to severe COVID-19, and how much by direct effects of prescribed drugs.  A secondary objective was to determine how much of the association could be explained by a sparse set of drug classes or other health-related factors, and how much was attributable to multifactorial causes, possibly including interactions between these factors.

@@

## Case ascertainment and sampling of controls 
All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  For each test-positive case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  Socioeconomic status was recorded as the quintile of the Scottish Index of Multiple Deprivation, based on the postal code.

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2.

British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.  The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).  BNF subparagraph codes generally represent pharmacological drug classes.  For this analysis prescription codes from chapters 14 and above, comprising mostly dressings and appliances but also vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each drug were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.  Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to the cutoff of 15 days before the date that the case (in each stratum of one case and ten matched controls) first tested positive.

# Statistical analysis 
Rate ratios for severe COVID-19 were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.  As described previously, we derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  For this analysis the list was extended to include non-cardiac circulatory disease as our analyses of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that condition is the only indication.  Diagnoses of diabetes were extracted from linkage to the national diabetes register.  The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material.

@@

We constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.  As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, a independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls. By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.  The propensity score was constructed by stepwise logistic regression with ever-use of a proton pump inhibitor as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription. As the case-control analysis is conditioned on age and sex, the terms for age and sex were dropped when the score was computed in the case-control dataset.

# Results
## Univariate associations of severe COVID-19 with drug classes
We first examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.  Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed for at least 10 individuals and were associated with severe disease at $p < 0.001$.  These COVID-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  This led us to examine associations with polypharmacy, indexed by the number of different drug classes for which prescriptions had been dispensed.

## Relation of severe COVID-19 to number of drug classes dispensed. 
Table \ref{tab:numdrugs.agegr} shows that severe COVID-19 was strongly related to the total number of different drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed since 1 June 2019.  The strength of this association was present in all age groups. With no dispensed prescriptions as reference category, the rate ratios associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.

@@

## Backward selection to identify specific drug classes contributing to the association of severe COVID-19 with number of drug classes dispensed
A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with number of drugs dispensed.  To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.  Dropping the first `r firstrow.neg - 1` drug classes from this subset increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped. Table \ref{tab:stepwise.subparas} shows how the log-likelihood decreased (though the rate ratio increased) as these variables were dropped from the model.

The only cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.  Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.  The last drug class retained by backward selection, and thus the variable with largest independent explanatory power, was proton pump inhibitors.

## Associations of severe COVID-19 with proton pump inhibitor use
This unexpected finding led us to investigate the association of COVID-19 with exposure to proton pump inhibitors in more depth.  This is challenging because proton pump inhibitors are usually co-prescribed with other drugs.  We used several approaches:

* Testing for a dose-response relationship

* Examination of the relationship by age.

* Adjusting for prespecified covariates: care home residence, SIMD quintile, index of hospital diagnosis of a disorder of the esophagus, stomach or duodenum, ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.

* Comparison in ever-exposed individuals between recent and less recent time windows of exposure.

* Stratification by propensity score for ever-use of a proton pump inhibitor.

For these analyses the dataset was not restricted to exclude care home residents or those  diagnosed with a listed condition.

### Dose-response relationship
Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inibitor, expressed as average DDDs/day over the period for which prescribing was observed. Table \ref{tab:ppinames} compares the effects of the five drugs in this class.  Omeprazole and lansoprazole were by far the most commonly prescribed drugs.  The effect size, expressed as rate ratio for unit increase in dose expressed as average DDDs/day, was higher for lansoprazole than for omeprazole.  This suggests that the DDDs for lansoprazole (30 mg) and omeprazole (20 mg) are not equivalent in potency.

### Relationship by age
Table \ref{tab:everuse.protonpump} shows the association with ever-use of proton pump inhibitors by age group, without and with adjustment for prespecified covariates.  The association was strongest in those aged under 60 years, and was reduced by adjusting for covariates.  Table \ref{tab:fatal.protonpump} shows the univariate associations of ever-use with fatal COVID-19 by age group.

Table \ref{tab:dosegr.protonpump} shows the rate ratio per unit increase in dose of proton pump inibitor, calculated as average defined daily doses dispensed per day during the observation period.  There is a strong dose-response relationship in those aged less than 75 years, but not in those aged over 75 years.

@@

### Construction of a propensity score
A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section. The `r nrow(coeffs.propensity)` variables retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.  As expected, the variables retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor. Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, the association of COVID-19 with use of proton pump inhibitors remained strong only in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.  <!--- show coefficient for COVID of each drug class in propensity score before and after adjustment for PPI? --->
